The hydra and the rooster

### 2020-07-23 latex/chapter-ks.tex
redaction introduction to ordinals

\begin{itemize}
\item There is no variant mapping the type \texttt{Hydra} into the interval $[0,\omega^2[$ (section ~\vref{omega2-case}), and a fortiori 
$[0,\omega[$

\item There is a variant that maps the type \texttt{Hydra} into the
interval $[0,\epsilon_0[$ (theorem \texttt{every\_battle\_terminates}, in section~\vref{hydra-variant}).
\end{itemize}

Thus, a very natural question is the following one:
\begin{quote}
  `` Is there  any variant from
\texttt{Hydra} into some interval $[0,\mu[$, where $\mu<\epsilon_0$, for proving the termination of all hydra battles ?''
\end{quote}

@@

\begin{quote}
There is no variant for proving the termination of all hydra battles
from \texttt{Hydra} into the interval $[0..\mu[$, where
$\mu< \epsilon_0$.
The same impossibility holds even if we consider only standard battles (with the successive replication factors $0,1,2,\dots,t,t+1,\dots$).
\end{quote}

@@

In order to apply our knowledge about  ordinal numbers (less than $\epsilon_0$) to the study of hydra battles, we define an injection
from the interval $[0,\epsilon_0[$ into the type \texttt{Hydra}.

@@

The next step of our development is to extend this relationship to
the order $<$ on $[0,\epsilon_0[$ on one side, and hydra fights on the other side.

>>>>

\begin{itemize}
\item There is no variant mapping the type \texttt{Hydra} into the interval $[0,\omega^2)$ (section ~\vref{omega2-case}), and a fortiori 
$[0,\omega)$

\item There is a variant that maps the type \texttt{Hydra} into the
interval $[0,\epsilon_0)$ (theorem \texttt{every\_battle\_terminates}, in section~\vref{hydra-variant}).
\end{itemize}

Thus, a very natural question is the following one:
\begin{quote}
  `` Is there  any variant from
\texttt{Hydra} into some interval $[0,\mu)$, where $\mu<\epsilon_0$, for proving the termination of all hydra battles ?''
\end{quote}

@@

\begin{quote}
There is no variant for proving the termination of all hydra battles
from \texttt{Hydra} into the interval $[0..\mu)$, where
$\mu< \epsilon_0$.
The same impossibility holds even if we consider only standard battles (with the successive replication factors $0,1,2,\dots,t,t+1,\dots$).
\end{quote}

@@

In order to apply our knowledge about  ordinal numbers (less than $\epsilon_0$) to the study of hydra battles, we define an injection
from the interval $[0,\epsilon_0)$ into the type \texttt{Hydra}.

@@

The next step of our development is to extend this relationship to
the order $<$ on $[0,\epsilon_0)$ on one side, and hydra fights on the other side.

### 2020-07-23 latex/chapter-large-sets.tex
redaction introduction to ordinals

Let us consider two integers $k$ and $l$, such that $0<k<l$. In order to check whether the interval $[k,l]$ is minimally large for $\alpha$, it is enough to
follow from $\alpha$ the path associated with the interval $[k,l[$ and verify that the last ordinal we obtain is equal to $1$.

@@

Let $0<\alpha<\epsilon_0$ be an ordinal term. We consider the functions which associate to each stritly positive integer $k$ the number $l$, where 
the interval $[k,l-1]$\;\footnote{also denoted by $[k,l[$}  is $\alpha$-mlarge.

>>>>

Let us consider two integers $k$ and $l$, such that $0<k<l$. In order to check whether the interval $[k,l]$ is minimally large for $\alpha$, it is enough to
follow from $\alpha$ the path associated with the interval $[k,l($ and verify that the last ordinal we obtain is equal to $1$.

@@

Let $0<\alpha<\epsilon_0$ be an ordinal term. We consider the functions which associate to each stritly positive integer $k$ the number $l$, where 
the interval $[k,l-1]$\;\footnote{also denoted by $[k,l)$}  is $\alpha$-mlarge.

### 2020-07-23 latex/chapter-ordinals.tex
redaction introduction to ordinals

\begin{figure}[h]
  \centering
\fbox{\Large
  \begin{minipage}{1.0\linewidth}
  \begin{align*}
     &\textcolor{blue}{0},\,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,\ldots\\
&\textcolor{red}{\omega},\,\omega+1,\omega+2,\omega+3,\ldots\\
&\textcolor{red}{\omega\times 2},\,\omega\times 2+1,\ldots, \textcolor{red}{\omega\times 3},\,\omega\times 3+1,\ldots, \textcolor{red}{\omega\times 4},\ldots,\\
&\textcolor{red}{\omega^2},\ldots, \textcolor{red}{\omega^2\times 42},\ldots,\textcolor{red}{\omega^3},\ldots, \textcolor{red}{\omega^4},\omega^4+1,\ldots,\\
&\textcolor{red}{\omega^\omega2},\ldots, \textcolor{red}{\omega^\omega\times 2},\ldots,\\
&\textcolor{red}{\omega^{\omega^\omega}},\ldots, \textcolor{red}{\omega^{\omega^\omega}+\omega^\omega\times 42+ \omega^{55}+\omega}, \ldots, \textcolor{red}{\omega^{\omega^{\omega+1}}}, \omega^{\omega^{\omega+1}}+1,\dots\\
& \textcolor{red}{\epsilon_0 (= \omega^\epsilon_0)}, \epsilon_0+1, \ldots,\\
& \textcolor{red}{\epsilon_1}, \ldots, \textcolor{red}{\epsilon_2}, \ldots, \textcolor{red}{\epsilon_\omega},\ldots \\
& \textcolor{red}{\Gamma_0}, \Gamma_0+1, \Gamma_0+2, \Gamma_0+3,\ldots, \textcolor{red}{\Gamma_0+\omega}, \ldots,\\
&\ldots
  \end{align*}   
  \end{minipage}}

@@

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, betaween $41$ and $43$.
\item Each ordinal in black is an immediate successor of anotter ordinal. We call it a \emph{successor} ordinal. For instance, $43$ is the successor of $42$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals that follow dots (written in red) are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). $\omega$ is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\end{itemize}

@@

The best known examples of well-ordered sets are the set $\mathbb{N}$ of natural numbers (with the usual order $<$), as well as any finite segment $[0,i[=\{j\in\mathbb{N}|j<i\}$.
The disjoint union of two copies of $\mathbb{N}$, \emph{i.e.} the set $\{0,1\}\times\mathbb{N}$ is also well-ordered,
with respect to the order below:

@@

If we consider an ordinal $\alpha$ as a well-ordered set, then its elements are just the ordinals strictly less than $\alpha$, \emph{i.e.} the \emph{segment} $\mathbb{O}_\alpha=[0, \alpha[$. So, we can speak about \emph{finite}, \emph{infinite}, \emph{countable}, etc., ordinals.

For instance, the elements of $\omega$ are the finite ordinals. $\omega$ is the first infinite ordinal.

@@

Chapter~\ref{chap:schutte} presents an adaptation to \coq{} of an axiomatization in classical logic of the set of countable ordinals by K. Schütte~\cite{schutte}. 
That formalization is quite complex and technical and unshamedly non-constructive,  so we put its description  in the last chapter of this document.

Please note that Schütte considers the (uncountable) set $\mathbb{O}$ of all countable ordinals. This set is well ordered (which is one of Schütte's axioms), and any ordinal $\alpha$ is associated with the segment $\mathbb{O}_\alpha$ 
of all ordinals strictly less than $\alpha$.

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as rather simple data types in \gallina. So, we will use \emph{ordinal notations}: Let $\alpha$ be some countable ordinal; In \coq{} terms, an ordinal notation for $\alpha$ will be a data type for representing all ordinals below $\alpha$ (excluding $\alpha$) and a bunch of functions for comparing ordinals, doing some arithmetic, inspecting the topology of the order (limits, successors), and proofs of well-foundedness). These functions should be proved correct, with respect to
Schütte's model (work in progress).

\section{Ordinal notations in Coq}

@@

Our definition of ordinal notation is inspired by our use of \coq{}: defining inductive data types, predicates, functions, etc, and proving properties of these functions.

@@

\item Let $\alpha$ be any ordinal, the \emph{segment} $\mathbb{O}_\alpha$ is the
interval $[0,\alpha[\,=\,\{\beta|0\leq\beta<\alpha\}$.

@@

\subsection{Finite ordinals}
Let $n$ be some natural number. The segment associated with $n$ is the interval 
$[0,n[\,=\,\{0,1,\dots,n-1\}$.

@@

For instance, let us build two elements of the segment $[0, 7[$, \emph{i.e.} two
inhabitants of   type (\texttt{t 7}), and prove a simple  inequality (see Fig.~\ref{fig:O7}).

@@

Finally, it is interesting to show that, if $i<j$, then the segment $[0,i[$ is a ``sub-segment'' of
$[0,j[$. Since the terms  $t\;i$ and $t\;j$ are not convertible, we consider a ``cast'' 
function $\iota$ from $t\;i$ to $t\;j$, and prove this function is  a bijection  from $t\;i$ to
the segment $[0,i[$ of $t\;j$.

@@

\begin{Coqsrc}
(** The segment associated with nA is isomorphic to
    the interval [0,b[ *)

>>>>

\begin{figure}[h]
  \centering
\fbox{\Large
  \begin{minipage}{1.0\linewidth}
  \begin{align*}
     &\textcolor{blue}{0},\,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,\ldots\\
&\textcolor{red}{\omega},\,\omega+1,\omega+2,\omega+3,\ldots\\
&\textcolor{red}{\omega\times 2},\,\omega\times 2+1,\ldots, \textcolor{red}{\omega\times 3},\,\omega\times 3+1,\ldots, \textcolor{red}{\omega\times 4},\ldots,\\
&\textcolor{red}{\omega^2},\ldots, \textcolor{red}{\omega^2\times 42},\ldots,\textcolor{red}{\omega^3},\ldots, \textcolor{red}{\omega^4},\omega^4+1,\ldots,\\
&\textcolor{red}{\omega^\omega},\ldots, \textcolor{red}{\omega^\omega+\omega^7\times 8},\ldots\textcolor{red}{\omega^\omega\times 2},\omega^\omega\times 2+1, \ldots,\\
&\textcolor{red}{\omega^{\omega^\omega}},\ldots, \textcolor{red}{\omega^{\omega^\omega}+\omega^\omega\times 42+ \omega^{55}+\omega}, \ldots, \textcolor{red}{\omega^{\omega^{\omega+1}}}, \omega^{\omega^{\omega+1}}+1,\dots\\
& \textcolor{red}{\epsilon_0 (= \omega^{\epsilon_0)}}, \epsilon_0+1, \epsilon_0+2, \epsilon_0+3, \ldots,\\
& \textcolor{red}{\epsilon_1}, \ldots, \textcolor{red}{\epsilon_2}, \ldots, \textcolor{red}{\epsilon_\omega},\ldots \\
& \textcolor{red}{\Gamma_0}, \Gamma_0+1, \Gamma_0+2, \Gamma_0+3,\ldots, \textcolor{red}{\Gamma_0+\omega}, \ldots,\\
&\ldots
  \end{align*}   
  \end{minipage}}

@@

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, betaween $41$ and $43$.
\item Each ordinal in black is an immediate successor of anotter ordinal. We call it a \emph{successor} ordinal. For instance, $43$ is the successor of $42$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals that  follow dots immediately (written in red) are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). $\omega$ is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item All the ordinals shown in this list are either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

The best known examples of well-ordered sets are the set $\mathbb{N}$ of natural numbers (with the usual order $<$), as well as any finite segment $[0,i)=\{j\in\mathbb{N}|j<i\}$.
The disjoint union of two copies of $\mathbb{N}$, \emph{i.e.} the set $\{0,1\}\times\mathbb{N}$ is also well-ordered,
with respect to the order below:

@@

In a set-theoretic framework, one can consider any ordinal $\alpha$ as a well-ordered set, whose  elements are just the ordinals strictly less than $\alpha$, \emph{i.e.} the \emph{segment} $\mathbb{O}_\alpha=[0, \alpha)$. So, one can speak about \emph{finite}, \emph{infinite}, \emph{countable}, etc., ordinals. Nevertheless, since we work with type theory, 
we do not identify ordinals as sets of ordinals, but the correspondance between ordinals and sets of ordinals is the function that maps $\alpha$ to $\mathbb{O}_\alpha$.
For instance $\mathbb{O}_\omega=\mathbb{N}$, and $\mathbb{O}_7=\{0,1,2,3,4,5,6\}$.

@@

\section{Ordinal numbers in Coq}

\subsection{Countable ordinals}

Chapter~\ref{chap:schutte} of this document presents an adaptation to \coq{} of an axiomatization in classical logic of the set of countable ordinals by K. Schütte~\cite{schutte}. 
That formalization is quite complex and technical and unshamedly non-constructive,  so we put its description  in the last chapter of this document.

Please note that Schütte considers the (uncountable) set $\mathbb{O}$ of all countable ordinals. This set is well ordered (which is one of Schütte's axioms), and associates to any ordinal $\alpha$ the segment $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.

In our adaptation to \coq{}, we declare a type \texttt{Ord}, a binary relation \texttt{lt} (with infix notation \texttt{``\_<\_''}, and assume Schütte's axiom. In Chapter~\ref{chap:schutte},
we derive some interesting properties of countable ordinals from these axioms.

\subsection{Ordinal notations}

Fortunately, the ordinals we need for  studying hydra battles are much simpler than Schütte's, and can be represented as rather simple data types in \gallina. So, we will use \emph{ordinal notations}: Let $\alpha$ be some countable ordinal; In \coq{} terms, an ordinal notation for $\alpha$ will be a data type for representing all ordinals below $\alpha$ (excluding $\alpha$) and a bunch of functions for comparing ordinals, doing some arithmetic, inspecting the topology of the order (limits, successors), and proofs of well-foundedness). These functions should be proved correct, with respect to Schütte's model.

@@

The following class definition, parameterized with a type $A$, a binary relation \texttt{lt} on $A$,

@@

\item Let $\alpha$ be any ordinal, the \emph{segment} $\mathbb{O}_\alpha$ is the
interval $[0,\alpha)\,=\,\{\beta|0\leq\beta<\alpha\}$.

@@

\subsection{Finite ordinals}
Let $n$ be some natural number. The segment associated with $n$ is the interval 
$)0,n[\,=\,\{0,1,\dots,n-1\}$.

@@

For instance, let us build two elements of the segment $[0, 7)$, \emph{i.e.} two
inhabitants of   type (\texttt{t 7}), and prove a simple  inequality (see Fig.~\ref{fig:O7}).

@@

Finally, it is interesting to show that, if $i<j$, then the segment $[0,i)$ is a ``sub-segment'' of
$[0,j)$. Since the terms  $t\;i$ and $t\;j$ are not convertible, we consider a ``cast'' 
function $\iota$ from $t\;i$ to $t\;j$, and prove this function is  a bijection  from $t\;i$ to
the segment $[0,i)$ of $t\;j$.

@@

\begin{Coqsrc}
(** The segment associated with nA is isomorphic to
    the interval [0,b) *)

### 2020-07-23 latex/chapter-schutte.tex
redaction introduction to ordinals

In the present chapter, we  compare our implementation of the segment $[0,\epsilon_0[$ with a mathematical text in order to ``validate'' our constructions.
We chose as reference the axiomatic definition of the set of countable ordinals,
in chapter V of Kurt Schütte's book `` Proof Theory ''~\cite{schutte}.

>>>>

In the present chapter, we  compare our implementation of the segment $[0,\epsilon_0)$ with a mathematical text in order to ``validate'' our constructions.
We chose as reference the axiomatic definition of the set of countable ordinals,
in chapter V of Kurt Schütte's book `` Proof Theory ''~\cite{schutte}.
